[Result, but poetic]: Does LLM Use Improve Data Science Education?

### 2024-08-02 paper/paper.qmd
Further cleaned up visualizations

# Create a multi-bar plot for coding self-perceptions grouped by response

# Select first three statements
writing_perceptions <- survey_data |>
  select(
    coding_is_easy_for_me,
    i_like_to_code,
    i_believe_it_is_important_to_be_a_good_coder
  )

writing_perceptions_long <- writing_perceptions |>
  pivot_longer(cols = everything(), names_to = "Statements", values_to = "Response") |>
  mutate(
    Statements = case_when(
      Statements == "coding_is_easy_for_me" ~ "\"Coding is easy for me\"",
      Statements == "i_like_to_code" ~ "\"I like to code\"",
      Statements == "i_believe_it_is_important_to_be_a_good_coder" ~ "\"I believe it is important\n to be a good coder\"",
      TRUE ~ Statements
    ),
    Response = case_when(
      Response == "This is a lot like me" ~ "A lot like me",
      Response == "This is very different to me" ~ "Very different from me",
      Response == "This somewhat describes me" ~ "Somewhat describes me",
    )
  ) |>
  filter(!is.na(Response))

# Taken from: https://stackoverflow.com/questions/52297978/decrease-overal-legend-size-elements-and-text
addSmallLegend <- function(myPlot, pointSize = 0.5, textSize = 3, spaceLegend = 0.1) {
  myPlot +
    guides(
      shape = guide_legend(override.aes = list(size = pointSize)),
      color = guide_legend(override.aes = list(size = pointSize))
    ) +
    theme(
      legend.title = element_text(size = textSize),
      legend.text = element_text(size = textSize),
      legend.key.size = unit(spaceLegend, "lines")
    )
}

@@

# Make the plot
multi_bar <- writing_perceptions_long |>
  ggplot(aes(x = Response, fill = Statements)) +
  geom_bar(position = "dodge", stat = "count") +
  labs(
    title = "Student Self-Perceptions of Coding",
    x = "Response",
    y = "Number of Students"
  ) +
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    axis.text.x = element_blank(),
  )

multi_bar <- addSmallLegend(multi_bar, pointSize = 0.5, textSize = 10, spaceLegend = 0.8)

@@

# Create a multi-bar plot for coding self-perceptions grouped by response

# Select first three statements
writing_perceptions <- survey_data |>
  select(
    when_i_check_my_code_it_is_easy_for_me_to_catch_my_mistakes,
    i_feel_confident_sharing_my_code,
    i_am_confident_in_my_overall_coding_ability
  )

writing_perceptions_long <- writing_perceptions |>
  pivot_longer(cols = everything(), names_to = "Statements", values_to = "Response") |>
  mutate(
    Statements = case_when(
      Statements == "when_i_check_my_code_it_is_easy_for_me_to_catch_my_mistakes" ~ "\"When I check my code, it is\n easy for me to catch my mistakes\"",
      Statements == "i_feel_confident_sharing_my_code" ~ "\"I feel confident sharing my code\"",
      Statements == "i_am_confident_in_my_overall_coding_ability" ~ "\"I am confident in\n my overall coding ability\"",
      TRUE ~ Statements
    ),
    Response = case_when(
      Response == "This is a lot like me" ~ "A lot like me",
      Response == "This is very different to me" ~ "Very different from me",
      Response == "This somewhat describes me" ~ "Somewhat describes me",
    )
  ) |>
  filter(!is.na(Response))

# Taken from: https://stackoverflow.com/questions/52297978/decrease-overal-legend-size-elements-and-text
addSmallLegend <- function(myPlot, pointSize = 0.5, textSize = 3, spaceLegend = 0.1) {
  myPlot +
    guides(
      shape = guide_legend(override.aes = list(size = pointSize)),
      color = guide_legend(override.aes = list(size = pointSize))
    ) +
    theme(
      legend.title = element_text(size = textSize),
      legend.text = element_text(size = textSize),
      legend.key.size = unit(spaceLegend, "lines")
    )
}

# Make the plot
multi_bar <- writing_perceptions_long |>
  ggplot(aes(x = Response, fill = Statements)) +
  geom_bar(position = "dodge", stat = "count") +
  labs(
    title = "Student Self-Perceptions of Coding",
    x = "Response",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
  )

@@

@@

ggplot(appropriate, aes(x = "", y = n, fill = ai_schoolwork_appropriate)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  labs(
    title = "Do students think AI use is appropriate in class?",
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  scale_fill_discrete(name = "Response") +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_blank(), # Remove x-axis text
    axis.title.x = element_blank(),
  ) +
  geom_text(aes(label = n),
    position = position_stack(vjust = 0.5)
  )
```

@@

To understand the role of AI chatbots in learning, students identified their usage by checking off various pre-defined use cases in the survey. As we can see in @fig-AI-usage-type-comp, technical questions and explaining concepts were the two top use cases among students. More than half of students also used ChatGPT for Quick Questions, General Knowledge, Writing Paper Code, and Checking Solutions. Just less than half used it to write paper content, which could suggest that students do not feel confident using it to improve writing, concurring with student attitudes in @Cahill_2024.

@@

@@

combined_data |>
  with(
    wordcloud(
      words = term,
      freq = n,
      scale = c(3, 0.5),
      min.freq = 1, # Adjust scale as needed
      max.words = 100,    # Maximum number of words to display
      colors = "blue"     # Color for the word cloud
    )
  )

@@

# Word cloud based on bigrams
word_counts |>
  with(
    wordcloud(
      words = word,
      freq = n,
      scale = c(3, 0.5),  # Adjust scale as needed
      max.words = 100,    # Maximum number of words to display
      colors = "blue"     # Color for the word cloud
    )
  )
```

>>>>

# Select first three statements for coding self-perceptions
coding_perceptions <- survey_data |>
  select(
    coding_is_easy_for_me,
    i_like_to_code,
    i_believe_it_is_important_to_be_a_good_coder
  )

# Convert to long format and adjust labels
coding_perceptions_long <- coding_perceptions |>
  pivot_longer(cols = everything(), names_to = "Statements", values_to = "Response") |>
  mutate(
    Statements = case_when(
      Statements == "coding_is_easy_for_me" ~ "\"Coding is easy for me\"",
      Statements == "i_like_to_code" ~ "\"I like to code\"",
      Statements == "i_believe_it_is_important_to_be_a_good_coder" ~ "\"I believe it is important\n to be a good coder\"",
      TRUE ~ Statements
    ),
    Response = case_when(
      Response == "This is a lot like me" ~ "A lot like me",
      Response == "This is very different to me" ~ "Very different from me",
      Response == "This somewhat describes me" ~ "Somewhat describes me",
      TRUE ~ Response
    )
  ) |>
  filter(!is.na(Response))

@@

# Function to adjust legend size
addSmallLegend <- function(myPlot, pointSize = 0.5, textSize = 3, spaceLegend = 0.1) {
  myPlot +
    guides(
      shape = guide_legend(override.aes = list(size = pointSize)),
      color = guide_legend(override.aes = list(size = pointSize)),
      fill = guide_legend(title = "Response", override.aes = list(size = pointSize))
    ) +
    theme(
      legend.title = element_text(size = textSize),
      legend.text = element_text(size = textSize),
      legend.key.size = unit(spaceLegend, "lines")
    )
}

# Make the plot
multi_bar <- coding_perceptions_long |>
  ggplot(aes(x = Response, fill = Response)) +
  geom_bar(position = "dodge", stat = "count") +
  facet_wrap(~Statements) +
  labs(
    title = "Student Self-Perceptions of Coding",
    subtitle = "Responses to statements about coding ease, enjoyment, and importance",
    x = "Response",
    y = "Number of Students"
  ) +
  scale_fill_manual(values = response_colors, name = "Response") + # Using custom colors
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    axis.text.x = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )

# Display the plot
multi_bar
```

@@

# Select the statements for coding self-perceptions
coding_perceptions <- survey_data |>
  select(
    when_i_check_my_code_it_is_easy_for_me_to_catch_my_mistakes,
    i_feel_confident_sharing_my_code,
    i_am_confident_in_my_overall_coding_ability
  )

# Convert to long format and adjust labels
coding_perceptions_long <- coding_perceptions |>
  pivot_longer(cols = everything(), names_to = "Statements", values_to = "Response") |>
  mutate(
    Statements = case_when(
      Statements == "when_i_check_my_code_it_is_easy_for_me_to_catch_my_mistakes" ~ "\"When I check my code, it is\n easy for me to catch my mistakes\"",
      Statements == "i_feel_confident_sharing_my_code" ~ "\"I feel confident\n sharing my code\"",
      Statements == "i_am_confident_in_my_overall_coding_ability" ~ "\"I am confident in\n my overall coding ability\"",
      TRUE ~ Statements
    ),
    Response = case_when(
      Response == "This is a lot like me" ~ "A lot like me",
      Response == "This is very different to me" ~ "Very different from me",
      Response == "This somewhat describes me" ~ "Somewhat describes me",
      TRUE ~ Response
    )
  ) |>
  filter(!is.na(Response))

# Function to adjust legend size
addSmallLegend <- function(myPlot, pointSize = 0.5, textSize = 3, spaceLegend = 0.1) {
  myPlot +
    guides(
      shape = guide_legend(override.aes = list(size = pointSize)),
      color = guide_legend(override.aes = list(size = pointSize)),
      fill = guide_legend(title = "Response", override.aes = list(size = pointSize))
    ) +
    theme(
      legend.title = element_text(size = textSize),
      legend.text = element_text(size = textSize),
      legend.key.size = unit(spaceLegend, "lines")
    )
}

# Make the plot
multi_bar <- coding_perceptions_long |>
  ggplot(aes(x = Response, fill = Response)) +
  geom_bar(position = "dodge", stat = "count") +
  facet_wrap(~Statements) +
  labs(
    title = "Student Self-Perceptions of Coding",
    subtitle = "Responses to statements about coding confidence and ease",
    x = "Response",
    y = "Number of Students"
  ) +
  scale_fill_manual(values = response_colors, name = "Response") + # Using custom colors
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    axis.text.x = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )

@@

# Display the plot
multi_bar
```

@@

ggplot(appropriate, aes(x = "", y = n, fill = ai_schoolwork_appropriate)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  labs(
    title = "Do students think AI use is appropriate in class?",
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  scale_fill_manual(
      name = "Response", 
      values = c("lightgreen", "tomato", "lightblue")) +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_blank(), # Remove x-axis text
    axis.title.x = element_blank(),
  ) +
  geom_text(aes(label = n),
    position = position_stack(vjust = 0.5)
  )
```

### Task-Specific Helpfulness

@@

To understand the role of AI chatbots in learning, students identified their usage by checking off various pre-defined use cases in the survey. @fig-AI-usage-type-comp shows that technical questions and explaining concepts were the two top use cases among students. More than half of students also used ChatGPT for Quick Questions, General Knowledge, Writing Paper Code, and Checking Solutions. Just less than half used it to write paper content, which could suggest that students do not feel confident using it to improve writing, concurring with student attitudes expressed in @Cahill_2024.

@@

Based on the word cloud and the relative counts of

@@

combined_data |>
  with(
    wordcloud(
      words = term,
      freq = n,
      scale = c(3, 0.5),
      min.freq = 1, # Min number of occurences
      max.words = 100,    # Maximum number of words to display
      colors = "skyblue4"     # Color for the word cloud
    )
  )

@@

# Word cloud based on bigrams
word_counts |>
  with(
    wordcloud(
      words = word,
      freq = n,
      scale = c(3, 0.5),
      min.freq = 1,# Adjust scale as needed
      max.words = 100,    # Maximum number of words to display
      colors = "blue"     # Color for the word cloud
    )
  )
```
